Implications of taxonomic bias for microbial differential-abundance analysis

### 2022-08-19 abundance-measurement.Rmd
Make edits from AW and BC

This section extends the theoretical results of @mclaren2019cons to describe the effect that consistent taxonomic bias within an MGS experiment affects the relative and absolute abundances measured for various microbial species.
We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.

## Model of MGS measurement

@@

We distinguish between two types of species-level *relative abundances* within a sample.
The *proportion* $P_{i}^{(a)}$ of species $i$ in sample $a$ equals its abundance relative to the total abundance of all species in $S$,
\begin{align}
  (\#eq:prop)
  P_{i}^{(a)} &\equiv \frac{A_i^{(a)}}{A_\tot^{(a)}}.
\end{align}
The *ratio* $R_{i/j}^{(a)}$ between two species $i$ and $j$ equals the abundance of $i$ divided by that of $j$,
\begin{align}
  (\#eq:ratio)
  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
\end{align}

The measured proportion of a species is given by its proportion of all the assigned reads in a sample,
\begin{align}
  (\#eq:prop-meas)
  \tilde P_{i}^{(a)} &\equiv \frac{M_i^{(a)}}{M_\tot^{(a)}}.
\end{align}
We use the tilde to distinguish the measurement from the actual quantity being measured.
From Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that the measured and actual proportion are related by
\begin{align}
  (\#eq:prop-error)
  \tilde P_{i}^{(a)} &= P_{i}^{(a)} \cdot \frac{B_i}{\bar B^{(a)}}.
\end{align}
<!-- Taxonomic bias thus creates a fold-error (FE) in the measured proportion $\tilde P_{i}^{(a)}$ of species $i$ equal to the efficiency $B_i$ of species $i$ $divided by the mean efficiency $\bar B^{(a)}$ in the sample. -->
Taxonomic bias thus creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
Since the mean efficiency varies across samples, so does the FE.
This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \@ref(fig:error-proportions). 
Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE < 1) but over-measured (FE > 1) in Sample 2.
This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
A demonstration in bacterial mock communities is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

The measured ratio $\tilde R_{i/j}^{(a)}$ between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \tilde R_{i/j}^{(a)} \equiv \frac{M_{i}^{(a)}}{M_{j}^{(a)}}.
\end{align}
From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that the measured and actual ratio are related by
\begin{align}
  (\#eq:ratio-error)
  \tilde R_{i/j}^{(a)} = R_{i/j}^{(a)} \cdot \frac{B_i}{B_j}.
\end{align}
Taxonomic bias thus creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

@@

**Higher-order taxa:**
We can consider a higher-order taxon $I$, such as a genus or phylum, as a set of species, $\{i \in I\}$.
The abundance of taxon $I$ in sample $a$ is the sum of the abundances of its constituent species, $A_{I}^{(a)} \equiv \sum_{i \in I} A_{i}^{(a)}$.
Similarly, the read count of taxon $I$ is the sum $M_{I}^{(a)} \equiv \sum_{i \in I} M_{i}^{(a)}$.
We further define the efficiency of taxon $I$ as the abundance-weighted average of the efficiencies of its constituent species,
\begin{align}
  (\#eq:efficiency-general)
  B_I^{(a)} \equiv \frac{\sum_{i\in I} A_{i}^{(a)} B_{i}}{\sum_{i\in I} A_{i}^{(a)}}.
\end{align}
With these definitions, the read count for taxon $I$ can be expressed as
$M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}$.
Thus $B_I^{(a)}$ plays a role analogous to the efficiency of an individual species, but differs in that it need not be constant across samples:
If the constituent species have different efficiencies, then the efficiency of the higher-order taxon $I$ depends on the relative abundances of its constituents and so will tend to vary across samples (@mclaren2019cons).
As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent, higher-order taxa efficiencies $B_I^{(a)}$ and $B_J^{(a)}$ are used. 
In this way, we see that both proportions and ratios among higher-order taxa may have inconsistent FEs.

@@

Several extensions of the standard MGS experiment make it possible to measure absolute species abundances.
These extensions fall into two general approaches.
The first approach leverages information about the abundance of the total community; for example, @vandeputte2017quan measured total-community abundance using flow cytometry and multiplied this number by MGS genus proportions to obtain the absolute abundances of individual genera (@vandeputte2017quan).
A second approach leverages information about the abundance of one or more individual species; for example, a researcher might 'spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (@harrison2021theq).
We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.

@@

Suppose that the total abundance of all species in the sample, $A_{\tot}^{(a)}$, has been measured by a non-MGS method, yielding a measurement $\tilde A_\tot^{(a)}$.
The absolute abundance of an individual species can be measured by multiplying the species' proportion from MGS by this total-abundance measurement,
\begin{align}
  (\#eq:density-prop-meas)
  \tilde A_i^{(a)} &= \tilde P_i^{(a)} \tilde A_\tot^{(a)}.
\end{align}
Total-abundance measurements recently used for this purpose include counting cells with microscopy (@lloyd2020evid) or flow cytometry (@props2017abso, @vandeputte2017quan, @galazzo2020howt), measuring the concentration of a marker-gene with qPCR or ddPCR (@zhang2017soil, @barlow2020aqau, @galazzo2020howt, @tettamantiboshier2020comp), and measuring bulk DNA concentration with a florescence-based DNA quantification method (@contijoch2019gutm).

Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias.
Flow cytometry may, for example, yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency (@lloyd2013meta).
We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
Suppose that each species $i$ has an _absolute efficiency_ $B_{i}^{\mtot}$ for the total-abundance measurement that is constant across samples.
Further, let $\bar B^{\mtot (a)}$ be the abundance-weighted average of these efficiencies in sample $a$---that is, the mean efficiency of the total-abundance measurement.
Neglecting other error sources, the total-abundance measurement equals
\begin{align}
  (\#eq:total-density-error)
  \tilde A_\tot^{(a)} 
  &= \sum_{i\in S} A_i^{(a)} B_{i}^{\mtot}
  \\&= A_\tot^{(a)} \bar B^{\mtot (a)}.
\end{align}
<!-- Note: We have assumed that only species in S contribute to the total abundance measurement. -->

Species abundance measurements derived by this method are affected by taxonomic bias in both the MGS and total-abundance measurement.
We can determine the resulting fold error (FE) by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
\begin{align}
  (\#eq:density-prop-error)
  \tilde A_\tot^{(a)}
  = A_\tot^{(a)} \cdot \frac{B_i \bar B^{\mtot (a)}}{\bar B^{(a)}}.
\end{align}
Equation \@ref(eq:density-prop-error) indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will vary across samples unless the two perfectly covary.

>>>>

This section extends the theoretical results of @mclaren2019cons to describe how taxonomic bias in an MGS experiment affects the relative and absolute abundances measured for various microbial species.
We show that some approaches to quantifying species abundance yield constant fold errors (FEs), while others yield FEs that depend on overall community composition and thus can vary across samples.

## A model of MGS measurements

@@

We distinguish between two types of species-level *relative abundances* within a sample.
The *proportion* $P_{i}^{(a)}$ of species $i$ in sample $a$ equals its abundance divided by the total abundance of all species in $S$,
\begin{align}
  (\#eq:prop)
  P_{i}^{(a)} &\equiv \frac{A_i^{(a)}}{A_\tot^{(a)}}.
\end{align}
The *ratio* $R_{i/j}^{(a)}$ between two species $i$ and $j$ equals the abundance of $i$ divided by that of $j$,
\begin{align}
  (\#eq:ratio)
  R_{i/j}^{(a)} \equiv \frac{A_i^{(a)}}{A_j^{(a)}}.
\end{align}

The measured proportion of a species is given by its proportion of all the assigned reads in a sample,
\begin{align}
  (\#eq:prop-meas)
  \tilde P_{i}^{(a)} &\equiv \frac{M_i^{(a)}}{M_\tot^{(a)}}.
\end{align}
We use the tilde to distinguish the measurement from the actual quantity being measured.
From Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that the measured and actual proportion are related by
\begin{align}
  (\#eq:prop-error)
  \tilde P_{i}^{(a)} &= P_{i}^{(a)} \cdot \frac{B_i}{\bar B^{(a)}}.
\end{align}
Taxonomic bias creates a fold-error (FE) in the measured proportion of a species that is equal to its efficiency divided by the mean efficiency in the sample.
Since the mean efficiency varies across samples, so does the FE.
This phenomenon can be seen for Species 3 in the two hypothetical communities in Figure \@ref(fig:error-proportions). 
Species 3, which has an efficiency of 6, is under-measured in Sample 1 (FE < 1) but over-measured (FE > 1) in Sample 2.
This difference occurs because the even distribution of species Sample 1 yields a mean efficiency of 8.33; in contrast, the lopsided distribution in Sample 2, which is dominated by the low-efficiency Species 1, has a mean efficiency of just 3.15.
A demonstration in bacterial mock communities is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

The measured ratio $\tilde R_{i/j}^{(a)}$ between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \tilde R_{i/j}^{(a)} \equiv \frac{M_{i}^{(a)}}{M_{j}^{(a)}}.
\end{align}
From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that the measured and actual ratio are related by
\begin{align}
  (\#eq:ratio-error)
  \tilde R_{i/j}^{(a)} = R_{i/j}^{(a)} \cdot \frac{B_i}{B_j}.
\end{align}
Taxonomic bias creates a FE in the measured ratio that is equal to the ratio in the species' efficiencies; the FE is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-measured by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

@@

**Higher-order taxa:**
We can consider a higher-order taxon $I$, such as a genus or phylum, as a set of species, $\{i \in I\}$.
The abundance of taxon $I$ in sample $a$ is the sum of the abundances of its constituent species, $A_{I}^{(a)} \equiv \sum_{i \in I} A_{i}^{(a)}$.
Similarly, the read count of taxon $I$ is the sum $M_{I}^{(a)} \equiv \sum_{i \in I} M_{i}^{(a)}$.
We further define the efficiency of taxon $I$ as the abundance-weighted average of the efficiencies of its constituent species,
\begin{align}
  (\#eq:efficiency-general)
  B_I^{(a)} \equiv \frac{\sum_{i\in I} A_{i}^{(a)} B_{i}}{\sum_{i\in I} A_{i}^{(a)}}.
\end{align}
With these definitions, the read count for higher-order taxon $I$ can be expressed as
$M_{I}^{(a)} = A_{I}^{(a)} B_I^{(a)} F^{(a)}$.
Thus $B_I^{(a)}$ plays a role analogous to the efficiency of an individual species, but differs in that it is not constant across samples:
If the constituent species have different efficiencies, then the efficiency of the higher-order taxon $I$ depends on the relative abundances of its constituents and so will vary across samples (@mclaren2019cons).
As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent, higher-order taxa efficiencies $B_I^{(a)}$ and $B_J^{(a)}$ are used. 
In this way, we see that both proportions and ratios among higher-order taxa may have inconsistent FEs.

@@

Several extensions of the standard MGS experiment make it possible to measure absolute species abundances.
These extensions fall into two general approaches.
The first approach leverages information about the abundance of the total community; for example, @vandeputte2017quan measured total-community abundance using flow cytometry and multiplied this number by genus proportions measured by MGS to quantify the absolute abundances of individual genera (@vandeputte2017quan).
A second approach leverages information about the abundance of one or more individual species; for example, a researcher might 'spike in' a known, fixed amount of an extraneous species to all samples prior to MGS, and normalize the read counts of all species to the spike-in species (@harrison2021theq).
We consider each approach in detail to determine how taxonomic bias affects the resulting absolute-abundance measurements.

@@

Suppose that the total abundance of all species in the sample, $A_{\tot}^{(a)}$, has been measured by a non-MGS method, yielding a measurement $\tilde A_\tot^{(a)}$.
The absolute abundance of an individual species can be quantified by multiplying the species' proportion from MGS by this total-abundance measurement,
\begin{align}
  (\#eq:density-prop-meas)
  \tilde A_i^{(a)} &= \tilde P_i^{(a)} \tilde A_\tot^{(a)}.
\end{align}
Total-abundance measurements recently used for this purpose include counting cells with microscopy (@lloyd2020evid) or flow cytometry (@props2017abso, @vandeputte2017quan, @galazzo2020howt), measuring the concentration of a marker-gene with qPCR or ddPCR (@zhang2017soil, @barlow2020aqau, @galazzo2020howt, @tettamantiboshier2020comp), and measuring bulk DNA concentration with a florescence-based DNA quantification method (@contijoch2019gutm).

Importantly, these methods of measuring total abundance are themselves subject to taxonomic bias that is analogous to, but quantitatively different from, the MGS relative abundance measurements.
Flow cytometry may yield lower cell counts for species whose cells tend to clump together or are prone to lysis during steps involved in sample collection, storage, and preparation.
Marker-gene concentrations measured by qPCR are affected by variation among species in extraction efficiency, marker-gene copy number, and PCR binding and amplification efficiency (@lloyd2013meta).
We can easily understand the impact of taxonomic bias on total-abundance measurement under simplifying assumptions analogous to those in our MGS model.
Suppose that each species $i$ has an _absolute efficiency_ $B_{i}^{\mtot}$ for the total-abundance measurement that is constant across samples.
Further, let $\bar B^{\mtot (a)}$ be the abundance-weighted average of these efficiencies in sample $a$---that is, the mean efficiency of the total-abundance measurement.
Neglecting other error sources, the total-abundance measurement equals
\begin{align}
  (\#eq:total-density-error)
  \tilde A_\tot^{(a)} 
  &= \sum_{i\in S} A_i^{(a)} B_{i}^{\mtot}
  \\&= A_\tot^{(a)} \bar B^{\mtot (a)}.
\end{align}
<!-- Note: We have assumed that only species in S contribute to the total abundance measurement. -->

Species abundance measurements derived by this method (Equation \@ref(eq:density-prop-meas)) are affected by taxonomic bias in both the MGS and total-abundance measurement.
We can determine the resulting fold error (FE) in the estimate $\tilde A_i^{(a)}$ by substituting Equations \@ref(eq:prop-error) and \@ref(eq:total-density-error) into Equation \@ref(eq:density-prop-meas), yielding
\begin{align}
  (\#eq:density-prop-error)
  \tilde A_\tot^{(a)}
  = A_\tot^{(a)} \cdot \frac{B_i \bar B^{\mtot (a)}}{\bar B^{(a)}}.
\end{align}
Equation \@ref(eq:density-prop-error) indicates that the FE in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
As in the case of proportions (Equation \@ref(eq:prop-error)), the FE depends on sample composition through the two mean efficiency terms and so will, in general, vary across samples.

### 2022-08-19 case-studies.Rmd
Make edits from AW and BC

The taxonomic bias species present in 'mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (@mclaren2019cons).
In practice it is difficult to construct control communities that span the many species present in complex natural communities.
However, gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.

@@

@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gained by a species from being allowed to colonize first.
Both analyses were based on fold differences (FDs) in species proportions and so in principle were sensitive to taxonomic bias.
To improve accuracy, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.

We repeated the two DA analyses of @leopold2020host with and without calibration and found that the results did not meaningfully differ
([SI Analysis of host genetics and arrival order](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/),
[SI Analysis of priority effects](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/)).
To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities 
([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/), 
SI Figure \@ref(fig:leopold2020host-variation)).
Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3-fold and by just 1.5-fold between the two most dominant.
The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual species, it had a negligible impact on the inferred FDs and the DA analyses based on them.

We performed an additional DA analysis on the data from @leopold2020host to investigate whether any commensals increased in absolute concentration in response to infection ([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/#change-in-commensals-due-to-infection)).
The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5-fold to 5.2-fold.
Using gamma-Poisson (or negative-binomial) regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
Commensals typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
For commensal-host pairs with relatively small observed decreases, bias correction greatly reduced the magnitude of the negative LFDs or, in several cases, resulted in LFDs that were near 0 or slightly positive.

@@

To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project 
(@huttenhower2012stru, [SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/)).
On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; GM $\md$ GSD of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples (SI Figure \@ref(fig:gut)B; ratio of GSD in vaginal samples to GSD in stool samples of `r hmp_me_stats['iid', 'gm'] %>% round(1)` $\md$ `r hmp_me_stats['iid', 'gsd'] %>% round(1)` across 1000 simulations).
Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \@ref(fig:gut)C), which the GSD in the proportion of a gut species across gut samples an average of `r hmp_species_diff %>% round(2)`-fold lower than that of a vaginal species across vaginal samples.
Recall that the importance of bias estimation of the FD in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \@ref(differential-abundance)).
Thus these results suggest that although the mean efficiency likely varies less across stool than vaginal samples, bias may be just as problematic for proportion-based DA analyses.

@@

Yet we also observed cases where the mean efficiency varied substantially.
In several cases, the mean efficiency varied systematically with the condition of interest sufficient to cause substantial systematic error in LFD estimates (Scenario 3).
Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.
The mean efficiency cannot differ by more than the species proportions constituting it.
Therefore, the error in the estimated LFDs tend to be practically significant only for species whose LFDs are substantially smaller in magnitude than the largest LFD.

>>>>

The taxonomic bias species present in 'mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (@mclaren2019cons).
In practice it is difficult to construct control communities that span the many species present in complex natural communities.
However, gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since it is possible to assemble mock communities containing all species in known relative abundances.

@@

@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gained by a species from being allowed to colonize first.
Both analyses were based on fold differences (FDs) in species proportions and so in principle were sensitive to taxonomic bias.
To improve accuracy, the authors incorporated the bias measured from the control samples into analysis-specific calibration procedures.

We repeated the two DA analyses of @leopold2020host with and without calibration and found that the results did not meaningfully differ
([SI Analysis of host genetics and arrival order](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/),
[SI Analysis of priority effects](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/)).
To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities 
([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/), 
SI Figure \@ref(fig:leopold2020host-variation)).
Despite the 13-fold variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62 and a geometric standard deviation of 1.05.
This consistency in the mean efficiency was despite the fact that each species showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3-fold and by just 1.5-fold between the two most dominant.
The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual species, it had a negligible impact on the inferred FDs and the DA analyses based on them.

We performed an additional DA analysis on the data from @leopold2020host to investigate whether any commensals increased in absolute concentration in response to infection ([SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-08-leopold2020host-case-study/#change-in-commensals-due-to-infection)).
The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5-fold to 5.2-fold.
Using gamma-Poisson regression of the observed read counts, we estimated the average LFD in commensal species proportions following infection with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
The commensals' proportions typically decreased post-infection, which is expected given the pathogen's growth to most abundant community member and the sum-to-one constraint faced by proportions.
However, failure to account for bias caused the decrease to be overestimated, by an amount corresponding to the inverse change in log mean efficiency.
For commensal-host pairs with relatively small observed decreases, bias correction greatly reduced the magnitude of the negative LFDs or, in several cases, resulted in LFDs that were near 0 or slightly positive.

@@

To investigate this question, we analyzed bacterial species profiles of vaginal and stool samples derived from shotgun sequencing in the Human Microbiome Project 
(@huttenhower2012stru, [SI Analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-30-hmp-stool-vagina-comparison/)).
On average, stool profiles had substantially greater order-2 alpha diversity than vaginal profiles (SI Figure \@ref(fig:gut)A; geometric mean (GM) $\md$ geometric standard deviation (GSD) of `r hmp_div_stats['stool', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['stool', 'gsd'] %>% round(1)` in stool samples and `r hmp_div_stats['vagina', 'gm'] %>% round(1)` $\md$ `r hmp_div_stats['vagina', 'gsd'] %>% round(1)` in vaginal samples).
To assess the potential importance of bias for proportion-based DA analyses in the two ecosystems, we quantified the multiplicative variation in the mean efficiency across samples for a large number of possible taxonomic biases under the assumption that the measured profiles reflected the truth.
Across simulation replicates, the GSD in the mean efficiency was typically lower in stool than in vaginal samples (SI Figure \@ref(fig:gut)B; ratio of GSD in vaginal samples to GSD in stool samples of `r hmp_me_stats['iid', 'gm'] %>% round(1)` $\md$ `r hmp_me_stats['iid', 'gsd'] %>% round(1)` across 1000 simulations).
Notably, however, the multiplicative variation in species also tended to be lower (SI Figure \@ref(fig:gut)C); the GSD in the proportion of a gut species across gut samples was an average of `r hmp_species_diff %>% round(2)`-fold lower than that of a vaginal species across vaginal samples.
Recall that the importance of bias estimation of the FD in a species' proportion depends on GSD in the mean efficiency relative the GSD in the species' proportion (Section \@ref(differential-abundance)).
Thus these results suggest that although the mean efficiency likely varies less across stool than vaginal samples, bias may be just as problematic for proportion-based DA analyses.

@@

Yet we also observed cases where the mean efficiency varied substantially.
In several cases, the (log) mean efficiency co-varied sufficiently strongly with the condition of interest to cause substantial systematic errors in LFD estimates (Scenario 3).
Examples include the comparison of foliar fungal microbiomes pre- and post-infection, for which the mean efficiency substantially increased post-infection, and the comparison of vaginal microbiomes with low versus high diversity in the MOMS-PI study, for which the mean efficiency was typically lower in high-diversity samples.
Our analysis of marine sediment communities is consistent with a systematic decline of the mean efficiency with burial time that is sufficient to substantially inflate the estimated growth rates of slowly changing species.
The mean efficiency cannot differ by more than the species proportions constituting it.
Therefore, the error in the estimated LFDs tend to be practically significant only for species whose LFDs are substantially smaller in magnitude than the largest LFD.

### 2022-08-19 conclusion.Rmd
Make edits from AW and BC

Proportion-based DA methods encompass many of the most popular methods for analyzing relative and absolute abundances.
Importantly, however, we found that the degree and significance of the error due to bias depends on experimental context and may often be negligible.
In particular, if the mean efficiency is approximately constant across samples, then bias has a negligible effect on multiplicative and rank-based cross-sample comparisons, and DA results will be unaffected.
If the mean efficiency varies but is unassociated with the condition of interest, then bias merely serves to increase noise and does not create systematic errors in DA results.
It is only when the mean efficiency is associated with the condition being analyzed that large systematic errors can occur.
Our case studies suggest that this problematic scenario can occur but may well be the exception rather than the rule.
Thus our results are still consistent with taxonomic bias having had a limited impact on extant DA results.
Systematic investigation of how the mean efficiency affects DA results across a wide range of studies may increase our confidence in previous DA results and/or alert us to the conditions in which they are most suspect.

>>>>

Proportion-based DA methods encompass many of the most popular methods for analyzing relative and absolute abundances.
Importantly, however, if the mean efficiency is approximately constant across samples, then bias has a negligible effect on multiplicative and rank-based DA results.
If the mean efficiency varies but is unassociated with the condition of interest, then bias merely serves to increase noise and does not create systematic errors in DA results.
It is only when the mean efficiency is associated with the condition being analyzed that large systematic errors can occur.
Our case studies suggest that this problematic scenario does occur, but it may be the exception rather than the rule.
Systematic investigation of how the mean efficiency affects DA results across a wide range of studies may increase our confidence in previous DA results and/or alert us to the conditions in which they are most suspect.
